Computing likelihoods under the SMC for a general class of ARGs.

### 2024-07-10 paper.tex
Rework first paragraph

\begin{abstract}
Ancestral recombination graphs (ARGs) are the focus of much ongoing research
interest. Recent progress in inference has made ARG-based approaches feasible
across of range of applications, and many new methods using inferred ARGs as
input have appeared. This progress on the long-standing problem of ARG
inference has proceeded in two distinct directions. The first
class of method in which dramatic improvements have been made is in the
statistically rigorous inference of ARGs under the Sequentially Markov
Coalescent (SMC) model, which are now practical for tens-to-hundreds of
samples. The second class of recently-developed inference methods are based on
approximate models and heuristics which scale to samples sizes roughly three
orders of magnitude larger. Although these heuristic methods are reasonably
accurate under many metrics, one major problem is that the ARGs they 
estimate cannot easily be connected with population genetic theory.
Specifically, because the estimated ARGs lack detail about recombination
events, there is currently no way to compute their likelihood
under a population genetics model. In this paper we provide a 
backwards-time formulation of the SMC based on a total ordering of 
lineages, and show how this leads to straightforward description 
of the likelihood of a general class of ARG under the SMC. We show that this 
formulation does not require precise details of recombination events
to be estimated, and is robust to other topological issues such as 
polytomies. This opens up many future directions, most excitingly the possibility
of combining the strengths of both inference approaches. Applying
these ideas may allow us to 
make rigorous inference more scalable, and scalable inference 
more rigorous.
\end{abstract}

\section{Introduction}
% TODO need to rewrite this as its too tskit-centric. We don't need to 
% emphasise tskit/msprime so much here, it's not necessary and will 
% annoy people. 
Following recent breakthroughs in simulation and inference methods,
there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Applications using inferred ARGs as input have begun to
appear~\citep{
stern_approximate_2019,
osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024}.
Additionally, there is a rich suite of tools growing around
the \texttt{tskit} ARG library~\citep{ralph_efficiently_2020}
including simulation~\citep{kelleher_efficient_2016,kelleher_efficient_2018,
adrion_community_2020,terasaki_geonomics_2021,
baumdicker_efficient_2021,
korfmann_weak_2022, petr_slendr_2023,tagami_tstrait_2024},
ARG inference~\citep{kelleher_inferring_2019,speidel_method_2019,
wohns_unified_2022,mahmoudi_bayesian_2022,
rasmussen_espalier_2022,
zhang_biobank-scale_2023,zhan_towards_2023} and analysis
packages
% etc
\citep{nowbandegani_extremely_2023,fan_likelihood_2023}.
Infererence methods now tend to fall into two classes: those that
sample from a statistical
model~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020,mahmoudi_bayesian_2022,
deng_robust_2024}
or parsimony~\citep{ignatieva_kwarg_2021,rasmussen_espalier_2022},
and those based on
heuristics~\citep{kelleher_inferring_2019,speidel_method_2019,
zhang_biobank-scale_2023}.
Typically, the more rigorous approaches can handle 10-100 samples,
while the approximate methods can scale to 10s to hundreds of
thousands of samples. With the explosive growth in dataset
sizes in recent years~\citep[e.g.][]{bycroft_genome_2018,halldorsson_sequences_2022,
all_genomic_2024} there is a pressing need for more statistical
rigor in large scale ARG inference.

@@

% THE PROBLEM!!! OMG it's terrible!!!
Despite these recent advances in statistically rigorous ARG 
inference, however, there is still a gap of some three orders 
of magnitude between the sample-sizes that can be handled 
(hundreds) compared to the large-scale hueristic methods
like tsinfer and ARGneedle (hundreds of thousands).
[TODO working through a more positive narrative here, based on
the idea of combining the strengths of the approaches]
Although major breakthroughs in scale have been achieved by hueristic
ARG inference
methods~\citep{kelleher_inferring_2019,speidel_method_2019,zhang_biobank-scale_2023},
these methods are currently fundamentally incompatible with rigorous
inference under population genetic models such as the SMC. This is 
because these methods return approximate ARG
\emph{structures}~\citep{wong_general_2023} [WEAK, FIXME]. 
Specifically,
we cannot evaluate a likelihood of these ARGs in coalescent 
models under the classical 
\cite{kuhner_maximum_2000} formulation, as these formulations assume
that complete information about recombination events is present.
[NOTE the following is super weak, and needs a lot of work.]
This inability to connect the results of large-scale inference methods
with a generative model
is a fundamental limitation, and limits both the utility of 
current point estimates as well as the possibilities for further
inference. Specifically, we are unable to quantify uncertainty 
around heuristic point estimates. We cannot explore ARG space
around the point estimates using (e.g. MCMC), as we have no way definition
of improvement. We cannot compare models of demography, or evaluate the 
goodness of fit for model estimates.

@@

%% NOTE: this is some orphan text from the intro that can be rejiggered for the 
% discussion. We can talk about precision of recombination estimates a bit 
% here I think, as it is an important topic.

>>>>

\begin{abstract}
Ancestral recombination graphs (ARGs) are the focus of much ongoing research
interest. Recent progress in inference has made ARG-based approaches feasible
across of range of applications, and many new methods using inferred ARGs as
input have appeared. This progress on the long-standing problem of ARG
inference has proceeded in two distinct directions. The first
class of method in which dramatic improvements have been made is in the
statistically rigorous inference of ARGs under the Sequentially Markov
Coalescent (SMC) model, which are now practical for tens-to-hundreds of
samples. The second class of recently-developed inference methods are based on
approximate models and heuristics which scale to samples sizes roughly three
orders of magnitude larger. Although these heuristic methods are reasonably
accurate under many metrics, one major problem is that the ARGs they 
estimate cannot easily be connected with population genetic theory.
Specifically, because the estimated ARGs lack detail about recombination
events, there is currently no way to compute their likelihood
under a population genetics model. In this paper we provide a 
novel backwards-time formulation of the SMC 
and derive a straightforward definition of the likelihood
of the likelihood of a general class of ARG under the SMC. 
We show that this formulation does not require precise details of recombination events
to be estimated, and is robust to other topological inconveniences such as 
polytomies. This opens up many future directions, most excitingly the possibility
of combining the strengths of both inference approaches. Applying
these ideas may allow us to 
make rigorous inference more scalable, and scalable inference 
more rigorous.
\end{abstract}

\section{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Numerous applications taking advantage of this detailed history
in inferred ARGs are now appearing~\citep{
stern_approximate_2019,
osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024}
and it seems likely that many more will
follow~\citep{harris_database_2019,harris_using_2023}.
Three classes of ARG inference method have emerged as the focus of recent
work, and have all been successfully applied in different contexts. 
Parsimony
based methods~\citep{ignatieva_kwarg_2021,rasmussen_espalier_2022}
seek to minimise some aspect of the inferred ARG
(usually the number number of recombinations required), and work well for 
answering specific detailed questions about 
recombination~\citep[e.g.][]{ignatieva_ongoing_2022} when the expected
number of ancestral recombinations is small (tens to hundreds).
The second class of inference method is based on sampling ARGs statistically 
from a population genetics model~\citep{rasmussen_genome-wide_2014,
hubisz_mapping_2020,mahmoudi_bayesian_2022, deng_robust_2024}, and have 
been applied in several
contexts~\citep[e.g.][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
stankowski_genetic_2024}. These methods are well-suited to applications 
in which population genetics models are a good fit [TODO: Say this differently, 
I just want to distinguish from e.g. SARS-CoV-2 where pop gen models are 
basically meaningless]
in cases in which the expected number of recombinations is in the 
thousands to tens-of-thousands range (e.g., human datasets in which
we have X samples and a segment of Y Mb of chromosome [TODO quantify
from SINGER paper). 
The third class of inference method and based on heuristics
and approximate models and focus on large scale inference,
where the expected number of recombinations is in the millions or
tens of millions [TODO check numbers here]~\citep{kelleher_inferring_2019,
speidel_method_2019,zhang_biobank-scale_2023}.
Other methods specialised to particular applications have also 
recently been proposed~\citep{zhan_towards_2023,schaefer_ancestral_2021}.
See~\cite{wong_general_2023} for a detailed review of ARG inference 
methods.

Typically, the more rigorous approaches can handle 10-100 samples,
while the approximate methods can scale to 10s to hundreds of
thousands of samples. With the explosive growth in dataset
sizes in recent years~\citep[e.g.][]{bycroft_genome_2018,halldorsson_sequences_2022,
all_genomic_2024} there is a pressing need for more statistical
rigor in large scale ARG inference.

@@

% THE PROBLEM!!! OMG it's terrible!!!
Despite these recent advances in statistically rigorous ARG 
inference, however, there is still a gap of some three orders 
of magnitude between the sample-sizes that can be handled 
(hundreds) compared to the large-scale hueristic methods
like tsinfer and ARGneedle (hundreds of thousands).
[TODO working through a more positive narrative here, based on
the idea of combining the strengths of the approaches]
Although major breakthroughs in scale have been achieved by hueristic
ARG inference
methods~\citep{kelleher_inferring_2019,speidel_method_2019,zhang_biobank-scale_2023},
these methods are currently fundamentally incompatible with rigorous
inference under population genetic models such as the SMC. This is 
because these methods return approximate ARG
\emph{structures}~\citep{wong_general_2023} [WEAK, FIXME]. 
Specifically,
we cannot evaluate a likelihood of these ARGs in coalescent 
models under the classical 
\cite{kuhner_maximum_2000} formulation, as these formulations assume
that complete information about recombination events is present
[Also: \cite{guo_recombination-aware_2022}]
[NOTE the following is super weak, and needs a lot of work.]
This inability to connect the results of large-scale inference methods
with a generative model
is a fundamental limitation, and limits both the utility of 
current point estimates as well as the possibilities for further
inference. Specifically, we are unable to quantify uncertainty 
around heuristic point estimates. We cannot explore ARG space
around the point estimates using (e.g. MCMC), as we have no way definition
of improvement. We cannot compare models of demography, or evaluate the 
goodness of fit for model estimates.

@@

% Additionally, there is a rich suite of tools growing around
% the \texttt{tskit} ARG library~\citep{ralph_efficiently_2020}
% including simulation~\citep{kelleher_efficient_2016,kelleher_efficient_2018,
% adrion_community_2020,terasaki_geonomics_2021,
% baumdicker_efficient_2021,
% korfmann_weak_2022, petr_slendr_2023,tagami_tstrait_2024},
% ARG inference~\citep{kelleher_inferring_2019,speidel_method_2019,
% wohns_unified_2022,mahmoudi_bayesian_2022,
% rasmussen_espalier_2022,
% zhang_biobank-scale_2023,zhan_towards_2023} and analysis
% packages
% % etc
% \citep{nowbandegani_extremely_2023,fan_likelihood_2023}.
%% NOTE: this is some orphan text from the intro that can be rejiggered for the 
% discussion. We can talk about precision of recombination estimates a bit 
% here I think, as it is an important topic.
